_Score:_ 15.5/20
---
title: "Assignment 8"
output: html_document
---

__Student Name:__ Sasagawa Koki  
__Student ID:__ 999054646  

## Assignment 8: Gene Networks 2

This should be a knittable .Rmd file.  Include this file, a .pdf and a .html when you turn in the assignment.
```{r}
# be sure to change the path
cities <- read.table("./us_cities.txt", sep = "\t", header = TRUE)
rownames(cities) <- cities$X #make first column the rownames
cities <- cities[,-1] #remove first column
cities <- as.matrix(cities) #convert to matrix
cities # print matrix
```
Imagine the airline had only 1500 miles. Putting a constraint on our city networks lets create an adjacency matrix. 
```{r}
cities_mat <- cities < 1500 # leave original matrix intact
diag(cities_mat) <- 0 # we do not have to fly within each of cities :)
cities_mat # check out the adjacency matrix
```

**Exercise 1:**
Based on this 0 or 1 representation of our network, what city is the most highly connected? *Hint: sum the values down a column OR across a row for each city*

```{r}
#The number 1 indicates by row, 2 indicates by column
#The apply function allows us to apply a function to each row or column of the matrix 
apply(cities_mat, 1, sum)
```

> The city that is most connected is Chicago. 

What if you were to extend the range to 2000 miles in the above code. Does the highest connected city change? If so explain. 
```{r}
cities_mat2 <- cities < 2000 # leave original matrix intact
diag(cities_mat2) <- 0 # we do not have to fly within each of cities :)
cities_mat2 # check out the adjacency matrix

#calculate the sum of 1's 
apply(cities_mat2, 1, sum)
```

> The city that is most connected is Denver. Changing the limit of the range can change the result as it becomes less restrictive. The extended limit can now capture more cities that could possibly be connected.

```{r}
library(igraph) # load package
# make sure to use the 2000 mile distance cutoff 
cities_graph2 <- graph.adjacency(cities_mat2, mode = "undirected")
plot.igraph(cities_graph2)
```

**Exercise 2:**
What is the total number of nodes in the plot? 
What is the total number of edges in the plot?

> The total number of nodes in the plot is 9.
The total number of edges in the plot is 19.

*GMT* -1 If you use the equation you should get 20

**Exercise 3:**
Re-calculate the adjacency matrix with the cutoff value at 2300. Calculate the number of edges using the following code. What do you get?

```{r}
cities_mat3 <- cities < 2300 # leave original matrix intact
diag(cities_mat3) <- 0 # we do not have to fly within each of cities :)
cities_mat3 # check out the adjacency matrix

sum(cities_mat3)/2 # divide by 2 because the matrix has 2 values for each edge
```

> Using the new limit of 2300 miles, we get a total of 24 edges. 

**Exercise 4:**
Fill in the following 0 or 1 values for our gene network above.

**(Gene1)--(1)--(Gene2)**

**(Gene2)--(0)--(Gene3)**

*GMT* -0.5 this should also be a 1 value because you want absolute cutoff

**(Gene1)--(0)--(Gene3)**

**(Gene3)--(0)--(Gene4)**

```{r}
genes <- read.table("voom_transform_brassica.csv", sep = ",", header = TRUE)
#Check
head(genes)
genes <- genes[,-c(38,42,46)] # remove questionable library columns
DE_genes <- read.table("DEgenes_GxE.csv", sep = ",")
#Check
head(DE_genes)
DE_gene_names <- rownames(DE_genes)
GxE_counts <- as.data.frame(genes[DE_gene_names,])
genes_cor <- cor(t(GxE_counts)) # calculate the correlation between all gene pairs
```

**Exercise 5:**

**a** 
Create an adjacency matrix called `genes_adj85` for the genes use a cutoff of abs(correlation) > 0.85.  Remember to set the diagonal of the adjacency matrix to 0.  Create a second adjacency matrix `genes_adj95` using a cutoff of abs(correlation) > 0.95. See above code for cities.

```{r}
genes_adj85 <- genes_cor > 0.85 # leave original matrix intact
diag(genes_adj85) <- 0
head(genes_adj85) # check out the adjacency matrix

genes_adj95 <- genes_cor > 0.95
diag(genes_adj95) <- 0
head(genes_adj95) # check out the adjacency matrix
```

**b**
Now we can do some calculations. If our cutoff is 0.85, how many edges do we have in our 255 node network? What if we increase our cutoff to 0.95?

```{r}
#Calculating the number of edges
sum(genes_adj85)/2 # divide by 2 because the matrix has 2 values for each edge
sum(genes_adj95)/2
```

*GMT* -1 should have different values we want the abs of the correlation both positive and negative

> Using the .85 cut off we get 460 edges. Increasing the cut off drastically reduced the number of edges we get to 9. 

**Exercise 6:**
Use the following code to plot our networks using different thresholds for connectivity. What patterns do you see in the visualization of this data? *Note: the second plot will take a while to render*

```{r}
#Graph 0.85 cutoff
gene_graph85 <- graph.adjacency(genes_adj85, mode = "undirected") #convert adjacency to graph
comps <- clusters(gene_graph85)$membership                        #define gene cluster membership
colbar <- rainbow(max(comps)+1)                                   #define colors
V(gene_graph85)$color <- colbar[comps+1]                          #assign colors to nodes
plot(gene_graph85, layout = layout.fruchterman.reingold, vertex.size = 6, vertex.label = NA)

#Graph 0.95 cutoff
gene_graph95 <- graph.adjacency(genes_adj95, mode = "undirected") #convert adjacency to graph
comps2 <- clusters(gene_graph95)$membership                        #define gene cluster membership
colbar2 <- rainbow(max(comps2)+1)                                   #define colors
V(gene_graph95)$color <- colbar2[comps2+1]                          #assign colors to nodes
plot(gene_graph95, layout = layout.fruchterman.reingold, vertex.size = 6, vertex.label = NA)
```

*GMT* -1 What patterns do you see in the visualization of this data?

**Exercise 7:**
Use the following code to answer these two questions: In gene_graph85, what is the total graph density? In gene_graph85, what is the average path length? 

```{r}
graph.density(gene_graph85)
average.path.length(gene_graph85)
```

> The total graph density is 0.01420411
The average path length is 3.478598

*GMT* -1 data off because of not taking abs correlation


```{r}
gene_graph85 <- graph.adjacency(genes_adj85, mode = "undirected")
distMatrix <- shortest.paths(gene_graph85, v = V(gene_graph85), to = V(gene_graph85))
head(distMatrix)[,1:7]
```

```{r}
pl <- get.shortest.paths(gene_graph85, 2, 7)$vpath[[1]] # pull paths between node 2 and 7

V(gene_graph85)[pl]$color <- paste("green")          # define node color
E(gene_graph85)$color <- paste("grey")               # define default edge color
E(gene_graph85, path = pl)$color <- paste("blue")    # define edge color
E(gene_graph85, path = pl)$width <- 10               # define edge width
plot(gene_graph85, layout = layout.fruchterman.reingold, vertex.size = 6, vertex.label = NA)
```


**Exercise 8:**
Using what you know about graphs, repeat your analysis of the smaller cities matrix. Show your code to answer these questions.
What is the graph density of the cities network with a 1500 mile flight range?
```{R}
#Graph cities with a 1500 mile flight rang
cities_mat_graph <- graph.adjacency(cities_mat, mode = "undirected") #convert adjacency to graph
comps3 <- clusters(cities_mat_graph)$membership                        #define gene cluster membership
colbar3 <- rainbow(max(comps3)+1)                                   #define colors
V(cities_mat_graph)$color <- colbar3[comps3+1]                          #assign colors to nodes
plot(cities_mat_graph, layout = layout.fruchterman.reingold, vertex.size = 6, vertex.label = NA)
```

```{r}
graph.density(cities_mat_graph)
```

> The graph densities of the cities with a 1500 mile flight range is 0.4444444

What is the average path length of the cities network with a 1500 mile flight range?
```{r}
average.path.length(cities_mat_graph)
```

> The average path length of the cities network with a 1500 mile flight range is 1.888889

Find the shortest path between SEA and DC with 1500 mile flight range. Graph it.
```{r}
cities_mat_graph <- graph.adjacency(cities_mat, mode = "undirected")
cityDistMatrix <- shortest.paths(cities_mat_graph, v = V(cities_mat_graph), to = V(cities_mat_graph))
cityDistMatrix
```

```{r}
p2 <- get.shortest.paths(cities_mat_graph, "SEA", "DC")$vpath[[1]] # pull paths between SEA and DC

V(cities_mat_graph)[p2]$color <- paste("green")          # define node color
E(cities_mat_graph)$color <- paste("grey")               # define default edge color
E(cities_mat_graph, path = p2)$color <- paste("blue")    # define edge color
E(cities_mat_graph, path = p2)$width <- 10               # define edge width
plot(cities_mat_graph, layout = layout.fruchterman.reingold, vertex.size = 6)
```

> The shortest past from SEA to DC is SEA --> DEN --> CHI --> DC.

Find the shortest path between SEA and DC with 2300 mile flight range. Graph it.
```{r}
cities_mat3_graph <- graph.adjacency(cities_mat3, mode = "undirected")
cityDistMatrix2 <- shortest.paths(cities_mat3_graph, v = V(cities_mat3_graph), to = V(cities_mat3_graph))
cityDistMatrix2
```

```{r}
p3 <- get.shortest.paths(cities_mat3_graph, "SEA", "DC")$vpath[[1]] # pull paths between SEA and DC

V(cities_mat3_graph)[p3]$color <- paste("green")          # define node color
E(cities_mat3_graph)$color <- paste("grey")               # define default edge color
E(cities_mat3_graph, path = p3)$color <- paste("blue")    # define edge color
E(cities_mat3_graph, path = p3)$width <- 10               # define edge width
plot(cities_mat3_graph, layout = layout.fruchterman.reingold, vertex.size = 6)
```

> The shortest distance is SEA --> CHI --> DC. The route does not stop at DEN as the planes have more gas now, it is able to go directly from SEA to CHI. 
